---
title: "Update genotypes"
author: "Alexey Larionov"
date: "30 Jan 2021"
output:
  html_document:
    toc: true
    toc_float: true
editor_options:
  chunk_output_type: console
---

# Summary

Update the names of tables: remove prefix "selected_"

Make additional matrices:  

- Additively encoded genotypes  
- Alt fractions  

<style>
pre{
  overflow-x: auto;
}
pre code{
  word-wrap: normal;
  white-space: pre;
}
</style>

```{r echo=F}
options(width=999)
```

# Start section

```{r}

# Time
Sys.time()

# Memory
gc()

# Clean up
rm(list=ls())
graphics.off()

# Options
options(stringsAsFactors = F)

# Folders
base_folder <- "/Users/alexey/Documents"
project_folder <- file.path(base_folder,"wecare","final_analysis_2021","reanalysis_wo_danish_2021","s02_wes_wecare_nfe")
scripts_folder <- file.path(project_folder,"scripts","s05_update_and_cleaning")
setwd(scripts_folder)
data_folder <- file.path(project_folder,"data","s05_update_and_cleaning")
#dir.create(data_folder)

```

# Load data

```{r}

load(file.path(project_folder,"data","s04_select_samples","s02_select_samples.RData"))

# Update folders
base_folder <- "/Users/alexey/Documents"
project_folder <- file.path(base_folder,"wecare","final_analysis_2021","reanalysis_wo_danish_2021","s02_wes_wecare_nfe")
scripts_folder <- file.path(project_folder,"scripts","s05_update_and_cleaning")
data_folder <- file.path(project_folder,"data","s05_update_and_cleaning")

```

# Rename data

```{r}

ls()

# Rename
variants.df <- selected_variants.df
phenotypes.df <- selected_phenotypes.df

gt.mx <- selected_gt.mx

dp.mx <- selected_dp.mx
gq.mx <- selected_gq.mx
ad.mx <- selected_ad.mx

# Clean-up
rm(selected_variants.df, selected_phenotypes.df, selected_dp.mx,
   selected_gq.mx, selected_gt.mx, selected_ad.mx)

```

# Check data

```{r}

ls ()

# Check objects dimentions
dim(variants.df)
dim(phenotypes.df)

dim(gt.mx)

dim(dp.mx)
dim(gq.mx)
dim(ad.mx)

# Check that the tables and matrices are in sync
sum(phenotypes.df$wes_id != colnames(gt.mx))
sum(phenotypes.df$wes_id != colnames(gq.mx))
sum(phenotypes.df$wes_id != colnames(dp.mx))
sum(phenotypes.df$wes_id != colnames(ad.mx))

sum(variants.df$ID != rownames(gt.mx))
sum(variants.df$ID != rownames(gq.mx))
sum(variants.df$ID != rownames(dp.mx))
sum(variants.df$ID != rownames(ad.mx))

```

# Additive encoding of genotypes  

```{r}

# Check the original gt matrix
gt.mx[1:5,1:5]
unique(as.vector(gt.mx))
table(gt.mx, useNA="always")

# Make matrix of NA for additive genotyoes
genotypes.mx <- matrix(NA,nrow=nrow(gt.mx),ncol=ncol(gt.mx))
colnames(genotypes.mx) <- colnames(gt.mx)
rownames(genotypes.mx) <- rownames(gt.mx)

# Populate the matrix with additive genotypes
0 -> genotypes.mx[ gt.mx == "0/0" ]
#1 -> genotypes.mx[ gt.mx == "1/0" ]
1 -> genotypes.mx[ gt.mx == "0/1" ]
2 -> genotypes.mx[ gt.mx == "1/1" ]

# Check result
dim(genotypes.mx)
genotypes.mx[1:5,1:5]
unique(as.vector(genotypes.mx))
table(genotypes.mx, useNA="always")

# Clean-up
rm(gt.mx)

```

# Split ad (allelic depth) matrix

```{r}

# Check the source ad matrix
# (attention should be paid to NA-s and ./. issue, if present)
ad.mx[1:5,1:5]
sum(is.na(ad.mx))
sum(ad.mx==".,.", na.rm=T)
".,." -> ad.mx[is.na(ad.mx)]
sum(ad.mx==".,.", na.rm=T)

# Extract data from ad.mx to a vector
# (keep paying attention to the possible NA issue)
ad <- unlist(strsplit(ad.mx,","))
sum(is.na(ad))

# Extract ref and alt data to separate vectors
ref <- ad[seq(1,length(ad),2)]
alt <- ad[seq(2,length(ad),2)]

# Keep an eye to the possible NA ./. issue
sum(ref == ".")
sum(alt == ".")
NA -> ref[ref=="."]
NA -> alt[alt=="."]

# Make matrices with ref and alt
ref.mx <- matrix(as.integer(ref), nrow=nrow(ad.mx))
rownames(ref.mx) <- rownames(ad.mx)
colnames(ref.mx) <- colnames(ad.mx)

alt.mx <- matrix(as.integer(alt), nrow=nrow(ad.mx))
rownames(alt.mx) <- rownames(ad.mx)
colnames(alt.mx) <- colnames(ad.mx)

# Check results
genotypes.mx[1:5,1:5]
gq.mx[1:5,1:5]

ad.mx[1:5,1:5]
ref.mx[1:5,1:5]
alt.mx[1:5,1:5]

# Clean-up
rm(ad, ref, alt)

```

# Make Alt Fraction matrix

```{r}

# Calculate Alt Fraction matrix
sum.mx <- ref.mx + alt.mx
sum(sum.mx == 0, na.rm=T) # 0/0 = NaN
alt_fraction.mx <- alt.mx / sum.mx 
alt_fraction.mx[1:5,1:5]

# Change NaN to NA (NaN has been placed for 0/0)
sum(is.na(alt_fraction.mx))
sum(is.nan(alt_fraction.mx))
sum(ad.mx == "0,0")
8401611-8401419
NA -> alt_fraction.mx[is.nan(alt_fraction.mx)]
sum(is.na(alt_fraction.mx))
sum(is.nan(alt_fraction.mx))
alt_fraction.mx[1:5,1:5]

# Note: sum.mx is not always equal to dp.mx
sum(sum.mx != dp.mx, na.rm=T)/(nrow(dp.mx)*ncol(dp.mx))

# Clean-up
rm(sum.mx, ad.mx, ref.mx, alt.mx)

```

# Check data

```{r}

ls()

# Check objects dimentions
dim(genotypes.mx)
dim(phenotypes.df)
dim(variants.df)

dim(dp.mx)
dim(gq.mx)
dim(alt_fraction.mx)


# Check that the tables and matrices are in sync
sum(phenotypes.df$wes_id != colnames(genotypes.mx))
sum(phenotypes.df$wes_id != colnames(gq.mx))
sum(phenotypes.df$wes_id != colnames(dp.mx))
sum(phenotypes.df$wes_id != colnames(alt_fraction.mx))

sum(variants.df$ID != rownames(genotypes.mx))
sum(variants.df$ID != rownames(gq.mx))
sum(variants.df$ID != rownames(dp.mx))
sum(variants.df$ID != rownames(alt_fraction.mx))

```

# Save results

```{r}

save.image(file.path(data_folder,"s01_update_genotypes.RData"))

```

# Final section

```{r}

sessionInfo()
Sys.time()
gc()

```
